\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-16 src/020chap2.tex
Fixed a bug in chapter 2 (from last commit). Endnotes in chap 2 changed to footnotes. Tucker matrices coded. Tikz is going to be the way of the graphics in Infinite Jest.

% \begin{definition}{\emph{Matrices that define
%       graphs.\cite[Def.~2.4]{d08phd}}} %
\begin{definition}
  {\em Matrices that define graphs. \cite[Def.~2.4]{d08phd}.}  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
    \label{def::maxcliquematrix}
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\endnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\endnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\endnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in endnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs\endnote { \cite{mcc04} cites that the problem of
  recognizing interval graphs has significance in molecular
  biology. Interestingly, in the late 1950s, before the structure of
  DNA was well-understood, Seymour Benzer was able to show that the
  intersection graph of a large number of fragments of genetic
  material was an interval graph \cite{sb59}. This was regarded as
  compelling evidence that genetic information was somehow arranged
  inside a structure that had a linear topology which we now know to
  be true\tnote{linear arrangement of DNA/genes?}.} and circular-arc
graphs have a long history in research.  The interest around them is
due to their very desirable property that several problems that are
NP-hard \tnote[minor]{all NPC problems are NPH yes?} on general
graphs, like finding a maximum clique or minimum coloring or
independent set, are polynomial time solvable in these graph classes
\cite{clrs01}.
In a similar fashion, a lot of problems that are hard on general
matrices have efficient solutions on matrices with \COP or \CROP
\cite[more citations pg.\,33]{d08phd}.

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs\endnote{\COP has several other applications
  \tnote{fill up.}}, we will now see results that study recognition of \COP
in matrices in the following section.

\temptext{ tab~\ref{tab:graphmatrices} -  have an abridged version
  of table 2.1 in dom. see notes for the possibility of a
  ``circle diagram''.} 
\begin{table}[htbp]
  \centering
 \caption{\figtabsize Graph matrices \temptext{PLACEHOLDER}}
  \label{tab:graphmatrices}
\end{table}

%\clearpage      % Print pending floats

@@

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using the
result {\sc hmpv00}\tnote{cite hmpv00} \tnote{in endnote put the theorem 2.7 - dom
  pg 43} \cite{d08phd} which demonstrates how such a derivation can be
done. However, this does not necessarily yield an efficient
algorithm. We will see results that directly solve the problem on
matrices since it is known that questions~(\ref{q:testcop}) and
(\ref{q:copperm}) stated above for \COP are efficiently solvable.
Table~\ref{tab:cophistory} gives a snapshot of these results.

@@

\def\tabparboxw{10cm}
\begin{table}[htbp]
  \centering
  {
    \fontencoding{OT1}\fontfamily     {cmss}\selectfont
    % \sffamily
    \footnotesize
    % \includegraphics[scale=0.5]{../img/cophistory.png} % TEMPORARY

  \begin{tabular}{lll}
     \hline
     \hline
     % 
     1899 & \parbox[t]{\tabparboxw}{ First mention of \COP
       (archaeology)} & 
       \\ % \parbox[t]{2cm}{}

@@

     \rowcolor[gray]{0.8} %
     1992 & \parbox[t]{\tabparboxw}{ Linear
       time algorithm COT without \PQtree} & \cite{wlh02}\endnote{First
       published in \cite{wlh02isaac92}} \\

@@

The first polynomial time algorithm for \COP testing was by
\cite{fg65} which uses overlapping properties of columns with \un
s. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix\endnote{
  Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}} must
have \COP on rows.

@@

%\clearpage
\subsection{Tucker's forbidden submatrices for \COP}

@@

\cite{at72} discovered certain forbidden structures for convex
bipartite graphs\endnote{The terminology in \cite{at72} differs. It
  uses the term {\em graphs with $V_1$-consecutive arragement} instead
  of {\em convex bipartite graphs}.} and by definition of this graph
class, this translates to a set of forbidden submatrices for matrices
with \cop.  The following are the theorems from \cite{at72} that
acheived this characterization.

Theorem~\ref{th:tuckeratfree} states that convex bipartite graphs
cannot have {\em asteroidal triples}\endnote{If $G = (V,E)$ is a
  graph, a set of three vertices from $V$ form an {\em asteroidal
    triple} if between any two of them there exists a path in $G$ that
  does not contain any vertex from the closed neighborhood of the
  third vertex.} contained in the corresponding vertex
partition\endnote{The partition corresponds to columns (rows) if its
  half adjacency matrix has \COP columns (rows).}.
Theorem~\ref{th:tuckerforbidden} lists the structures in a
bipartite graph that force one of its vertex partitions to have
asteriodal triples -- in other words, it identifies the subgraphs that
prevent the graph from being convex bipartite.

\begin{theorem}
  (\cite[Th.~6]{at72}, \cite[Th.~2.3]{d08phd})\\
  A bipartite graph $G = (V_1, V_2, E)$ is convex bipartite on
  columns\endnote{Abridged to match terminology adopted in this
    document. See previous note.} \iff $V_1$ contains no asteroidal
  triple of $G$.
  \label{th:tuckeratfree}
\end{theorem}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.35]{../img/tuckersforbiddensubmatrices.png}  
  \caption[\figtabsize Tucker's forbidden submatrices]{\figtabsize Tucker's
    forbidden submatrices for convex bipartite graphs. \temptext{ PLACEHOLDER
    IMG }} 
  \label{fig:forbiddensubmatrices}
\end{figure}

@@

%\clearpage
\subsection{Booth and Lueker's $PQ$ tree -- a linear COT algorithm}

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm\endnote{Time complexity is $O\left(m+n+f\right)$ where $m
  \times n$ is the order of the input matrix and $f$ is the number of $\un$s in it.} for \cop testing while given a linear time interval graph
recognition algorithm by a simplification of \tnote{cite - A. Lempel,
  S. Even and I. Cederbaum, An Algorithm for Planarity Testing of
  Graphs, Theory of Graphs, ed., P. Rosenstiehl, Gordon and Breach,
  New York, (1967), 215-232.  }'s planarity test algorithm.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

\begin{definition}[\PQtree \cite{bl76, mcc04}]
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row)\endnote{Note
      that \COP order for column requires permutation of rows and vice
      versa.} for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}

@@

%\section{Certifying Algorithms}
\subsection{\PQRtree\ --  COP for set systems}
\label{sec:surveycertalgo}

Section~\ref{sec:motive} mentions how a binary matrix naturally maps
to a system of sets.  A set can be constructed for each column of
matrix with its elements being those row indices at which the column
has \un s. Thus the collection of sets corresponding each column of
the matrix forms a set system with universe as the set of all row
indices of the matrix.  This simple construction is formally described
in Definition~\ref{def:matrixsetsystem} along with the idea of \cop
for set systems\endnote{As seen in Section~\ref{sec:courseschedule}}.

\begin{definition}%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  [\emph{\Cop for set systems}.]  Let $M$ be a binary matrix of order
  $n \times m$ % , $r_i, i \in [n]$ be the rows in $M$
  and $\{c_i \mid i \in [m]\}$ be the columns in $M$. A set system
  $\cF_{M} = \{S_i \mid S_i \subseteq [n], i \in [m]\}$ is defined
  such that for every column $c_i$ of $M$, set $S_i = \{j \mid m_{ji}
  = \un \}$. The collection $\cF_M$ is the {\em set system of binary
    matrix} $M$.\\
  A set system $\cF$ from universe $U = \{1, 2, \ldots, n\}$ has the
  {\em \cop} if it is possible to assign a linear order, \ie a
  permutation $\sigma = u_1u_2\ldots u_n$ to $U$ \stt each set $S \in
  \cF$ appears as a consecutive subsequence $u_{i}u_{i+1}\ldots
  u_{i+k-1}$ (also called an interval) of $\lambda$ in the linear order, where
  positive integers $k = |S|$ and $i \le n+1-k$.
\end{definition}

@@

Generalized \PQtree or \gPQtree is a data structure defined in \cite{n89} to
represent all orthogonal sets of a set system $\cF$. {\em Orthogonal
  sets} of $\cF$ are sets that do not overlap with any set in $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system
$\cF$. \tnote{define substitutio decomp.} This is later used to define
a certificate if $\cF$ has no \COP. \PQRtree is a data structure
defined in \cite{mm96} as a data structure to represent any set system
$\cF$ with additional information in their $R$-nodes if $\cF$ has no
\COP.  All three of these data structures are proposed as
generalizations of \cite{bl76}'s \PQtree and hence produce the
$\PQtree$ if $\cF$ has \COP. As a whole, all these three data
strutures are near-identical with their differences being notional.
In this section, we will discuss the basic theory that they all hold
and stick to the name of the data structure as \PQRtree.

@@

An important observation made by \cite{mm96} is that if $\cF$ is a set
system with \COP then, aside from every set in $\cF$ being consecutive
(after applying the \COP order) the following sets must also be
consecutive for any $A, B \in \cF$.
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and $A \nsubseteq B$ respectively.
\end{enumerate}
Also note that trivially, sets in $\cT(U)$ are consecutive in any
permutation of $U$\endnote{$\emptyset$ is considered consecutive by
  convention.}.

@@

\tnote[FYI]{ The results in cite~{bl76} on COT are based on the
   result that interval graphs are AT-free chordal graphs cite~{lb62}?
 No they don't seem related. this is the stupid bl lb confusion.}%

>>>>

\begin{definition}
  {\em Matrices that define graphs. \cite[Def.~2.4]{d08phd}.}  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
    \label{def::maxcliquematrix}
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.
  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\footnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\footnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\footnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in footnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs\footnote{\cite{mcc04} cites that the problem of
  recognizing interval graphs has significance in molecular
  biology. Interestingly, in the late 1950s, before the structure of
  DNA was well-understood, Seymour Benzer was able to show that the
  intersection graph of a large number of fragments of genetic
  material was an interval graph \cite{sb59}. This was regarded as
  compelling evidence that genetic information was somehow arranged
  inside a structure that had a linear topology which we now know to
  be true from the discovery of linear structure of DNA.}and
circular-arc graphs have a long history in research.  The interest
around them is due to their very desirable property that several
problems that are NP-hard \tnote[minor]{all NPC problems are NPH yes?}
on general graphs, like finding a maximum clique or minimum coloring
or independent set, are polynomial time solvable in these graph
classes \cite{clrs01}.  In a similar fashion, a lot of problems that
are hard on general matrices have efficient solutions on matrices with
\COP or \CROP \cite[more citations pg.\,33]{d08phd}.

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs\footnote{\COP has several other applications
  \temptext{fill up.}}, we will now see results that study recognition of \COP
in matrices in the following section.

\temptext{ tab~\ref{tab:graphmatrices} -  have an abridged version
  of table 2.1 in dom. see notes for the possibility of a
  ``circle diagram''.} 
\begin{table}[htbp]
  \centering
% \includegraphics[scale=0.1]{../img/circlediagram_domtab2.1__2.png} %
 \caption{\figtabsize Graph matrices \temptext{PLACEHOLDER}}
  \label{tab:graphmatrices}
\end{table}

@@

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using the
result {\sc hmpv00}\tnote{cite hmpv00} \tnote{in footnote put the theorem 2.7 - dom
  pg 43} \cite{d08phd} which demonstrates how such a derivation can be
done. However, this does not necessarily yield an efficient
algorithm. We will see results that directly solve the problem on
matrices since it is known that questions~(\ref{q:testcop}) and
(\ref{q:copperm}) stated above for \COP are efficiently solvable.
Table~\ref{tab:cophistory} gives a snapshot of these results.

@@

\def\tabparboxw{10cm}
\begin{table}[htbp]
  \centering
  {
    \fontencoding{OT1}\fontfamily{cmss}\selectfont%
    \footnotesize%

    \begin{tabular}{lll}
      \hline
      \hline

      1899 & \parbox[t]{\tabparboxw}{ First mention of \COP
        (archaeology)} &                  \\

@@

      \rowcolor[gray]{0.8} %
      1992 & \parbox[t]{\tabparboxw}{ Linear
        time algorithm COT without \PQtree} & \cite{wlh02}%\footnote{First
      % published in \cite{wlh02isaac92}}
      \\

@@

The first polynomial time algorithm for \COP testing was by
\cite{fg65} which uses overlapping properties of columns with \un
s. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix\footnote{
  Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}} must
have \COP on rows.

@@

\subsection{Tucker's forbidden submatrices for \COP}

@@

\cite{at72} discovered certain forbidden structures for convex
bipartite graphs\footnote{The terminology in \cite{at72} differs. It
  uses the term {\em graphs with $V_1$-consecutive arragement} instead
  of {\em convex bipartite graphs}.} and by definition of this graph
class, this translates to a set of forbidden submatrices for matrices
with \cop.  The following are the theorems from \cite{at72} that
acheived this characterization.

Theorem~\ref{th:tuckeratfree} states that convex bipartite graphs
cannot have {\em asteroidal triples}\footnote{If $G = (V,E)$ is a
  graph, a set of three vertices from $V$ form an {\em asteroidal
    triple} if between any two of them there exists a path in $G$ that
  does not contain any vertex from the closed neighborhood of the
  third vertex.} contained in the corresponding vertex
partition\footnote{The partition corresponds to columns (rows) if its
  half adjacency matrix has \COP columns (rows).}.
Theorem~\ref{th:tuckerforbidden} lists the structures in a
bipartite graph that force one of its vertex partitions to have
asteriodal triples -- in other words, it identifies the subgraphs that
prevent the graph from being convex bipartite.

\begin{theorem}
  (\cite[Th.~6]{at72}, \cite[Th.~2.3]{d08phd})\\
  A bipartite graph $G = (V_1, V_2, E)$ is convex bipartite on
  columns\footnote{Abridged to match terminology adopted in this
    document. See previous note.} \iff $V_1$ contains no asteroidal
  triple of $G$.
  \label{th:tuckeratfree}
\end{theorem}

@@

\begin{figure}[htbp]
  \centering
%  \includegraphics[scale=0.35]{../img/tuckersforbiddensubmatrices.png}  
  \setlength\dashlinedash{1pt} %
  \setlength\dashlinegap{0pt}  %  Make non zero if cline can be done
                               %   with dashed lines.
  \singlespacing
  \begin{tabular}[h]{ll}
%   \small 
   \begin{tabular}[t]{|lll ll l|l} %                   MATRIX 1
      \multicolumn{7}{l}{\normalsize \hspace{-0.5in} $M_{I_k}, k \ge 1$}              \\   
      \multicolumn{6}{c}{\matrixbracetop{3.25}{$k+2$}}   \\
      \cline{1-6} % \firsthline %      Begin matrix  %%%%%%
      \un & \un & 0   & \multicolumn{2}{c}{\dots}  & 0    &
        \hspace{-0.1in}\multirow{5}{*}{\matrixbraceright{2.5}{$k+2$}}   \\    
      0   & \un & \un & 0 & ...& 0   \\
      \multicolumn{6}{|c|}{\dots}                        \\
      0   & \multicolumn{2}{c}{\dots}  & 0   & \un & \un \\
      \cline{1-6} % \hdashline %%%%%%%%%%%%%%%%%%%%%%%%%%%%
      \un & 0   & \multicolumn{2}{c}{\dots}  & 0   & \un \\
      \cline{1-6} % \lasthline %      End matrix  %%%%%%%%%
    \end{tabular}
    &%
%    \small
    \begin{tabular}[t]{|lll ll ll|l} %                   MATRIX 2
      \multicolumn{8}{l}{\normalsize \hspace{-0.5in}$M_{II_k}, k \ge 1$}                   \\
      \multicolumn{7}{c}{\matrixbracetop{3.8}{$k+3$}}            \\ 
      \cline{1-7}%  \firsthline %      Begin matrix  %%%%%%%%%%%%
      \un & \un & 0   & \multicolumn{3}{c:}{\dots}       & 0    &
        \hspace{-0.1in}\multirow{6}{*}{\matrixbraceright{3}{$k+3$}}         \\
      0   & \un & \un & 0 & \multicolumn{2}{c:}{\dots}   & 0   \\ 
      \multicolumn{6}{|c:}{\dots}                        &     \\
      0   & \multicolumn{2}{c}{\dots}  & 0   & \un              &
        \multicolumn{1}{l:}{\un} & 0                           \\
      \cline{1-6}    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      0   & \un & \multicolumn{4}{c}{\dots}              & \un \\
      \un & \multicolumn{3}{c}{\dots}        & \un & 0   & \un \\
      \cline{1-7}% \lasthline %      End matrix  %%%%%%%%%%%%%%%%
    \end{tabular}
    \\\\%
%    \small
    \begin{tabular}[t]{|lll ll l l|l} %                  MATRIX 3 
      \multicolumn{8}{l}{\normalsize \hspace{-0.5in}$M_{III_k}, k \ge 1$}                  \\
      \multicolumn{7}{c}{\matrixbracetop{3.8}{$k+3$}}            \\
      \cline{1-7}%      \firsthline %      Begin matrix %%%%%%%%%
      \un & \un & 0   & \multicolumn{3}{c:}{\dots}       & 0    &
        \hspace{-0.1in}\multirow{5}{*}{\matrixbraceright{2.5}{$k+2$}}         \\
      0   & \un & \un & 0 & \multicolumn{2}{c:}{\dots}   & 0   \\ 
      \multicolumn{6}{|c:}{\dots}                        &     \\
      0   & \multicolumn{2}{c}{\dots}  & 0   & \un              &
        \multicolumn{1}{l:}{\un}                         & 0   \\
      \cline{1-6}    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      0   & \un & \multicolumn{2}{c}{\dots}  & \un & 0   & \un \\
      \cline{1-7}%      \lasthline %      End matrix %%%%%%%%%%%%
    \end{tabular}
    &%
%    \small
    \begin{tabular}[t]{|llllll|}  %                     MATRIX 4
      \multicolumn{6}{l}{\normalsize \hspace{-0.5in}$M_{IV}$} \\
      \multicolumn{6}{l}{} \\
      \firsthline
      \un & \un & 0 & 0 & 0 & 0   \\
      0 & 0 & \un & \un & 0 & 0   \\
      0 & 0 & 0 & 0 & \un & \un   \\
      \un & 0 & \un & 0 & \un & 0 \\
      \lasthline
    \end{tabular}
    \\\\%
%    \small
    \begin{tabular}[t]{|lllll|}  %                     MATRIX 5
      \multicolumn{5}{l}{\normalsize \hspace{-0.5in}$M_{V}$} \\
      \multicolumn{5}{l}{} \\
      \firsthline
      \un & \un & 0   & 0   & 0   \\
      0   & 0   & \un & \un & 0   \\
      \un & \un & \un & \un & 0   \\
      \un & 0   & \un & 0   & \un \\
      \lasthline
    \end{tabular}
  \end{tabular}

  \caption[\figtabsize Tucker's forbidden submatrices]{\figtabsize Tucker's
    forbidden submatrices for convex bipartite graphs. \cite{at72}} 
  \label{fig:forbiddensubmatrices}
\end{figure}

@@

\subsection{Booth and Lueker's $PQ$ tree -- a linear COT algorithm}

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm\footnote{Time complexity is $O\left(m+n+f\right)$ where $m
  \times n$ is the order of the input matrix and $f$ is the number of
  $\un$s in it.} for \cop testing while given a linear time interval
graph recognition algorithm by a simplification of \tnote{cite -
  A. Lempel, S. Even and I. Cederbaum, An Algorithm for Planarity
  Testing of Graphs, Theory of Graphs, ed., P. Rosenstiehl, Gordon and
  Breach, New York, (1967), 215-232.  }'s planarity test algorithm.
% Hsu in wlh01 said BL76 gave a planarity test - "the induced PQ-tree
% algorithm can considerably simplify Booth and Lueker’s modification
% of Lempel, Even and Cederbaum’s planarity test. " 
%
% McConnell in hm03 said the first linear algo for interval graph
% recog was by BL76 - "Benzer’s work and other applications of
% interval graphs motivated a search for eﬃcient algorithms for
% recognizing interval graphs, and for constructing a set of intervals
% to represent the graph when it is one [8]. A linear-time algorithm
% was given in 1976 by Booth and Lueker [2]." 
\cite{bl76} introduces a data structure called \PQtree and their \COP
testing algorithm is a constructive one that outputs a \PQtree if the
input has \COP. A \PQtree represents all the \COP orderings of the
matrix it is associated with. \cite{bl76}'s algorithm uses the fact
that if a matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

%\begin{minipage}[h]{5.5in}
\begin{definition}[\PQtree \cite{bl76, mcc04}]
  A \PQtree of matrix $M$ with \COP on columns (rows), is a tree with
  the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row)\footnote{Note
      that \COP order for column requires permutation of rows and vice
      versa.} for $M$.
  \item Every non-leaf node in the tree is labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}
%\end{minipage}

@@

\subsection{\PQRtree\ --  COP for set systems}
\label{sec:surveycertalgo}

Section~\ref{sec:motive} mentions how a binary matrix naturally maps
to a system of sets.  A set can be constructed for each column of
matrix with its elements being those row indices at which the column
has \un s. Thus the collection of sets corresponding each column of
the matrix forms a set system with universe as the set of all row
indices of the matrix.  This simple construction is formally described
in Definition~\ref{def:matrixsetsystem} along with the idea of \cop
for set systems\footnote{As seen in Section~\ref{sec:courseschedule}}.

\begin{definition}%
  \label{def:matrixsetsystem}{\emph{\Cop for set systems}}%
  % [\emph{Set system of a binary matrix}.]
   Let $M$ be a binary matrix of order
  $n \times m$ % , $r_i, i \in [n]$ be the rows in $M$
  and $\{c_i \mid i \in [m]\}$ be the columns in $M$. A set system
  $\cF_{M} = \{S_i \mid S_i \subseteq [n], i \in [m]\}$ is defined
  such that for every column $c_i$ of $M$, set $S_i = \{j \mid m_{ji}
  = \un \}$. The collection $\cF_M$ is the {\em set system of binary
    matrix} $M$.\\
  A set system $\cF$ from universe $U = \{1, 2, \ldots, n\}$ has the
  {\em \cop} if it is possible to assign a linear order, \ie a
  permutation $\sigma = u_1u_2\ldots u_n$ to $U$ \stt each set $S \in
  \cF$ appears as a consecutive subsequence $u_{i}u_{i+1}\ldots
  u_{i+k-1}$ (also called an interval) of $\lambda$ in the linear order, where
  positive integers $k = |S|$ and $i \le n+1-k$.
\end{definition}

@@

Generalized \PQtree or \gPQtree is a data structure defined in \cite{n89} to
represent all orthogonal sets of a set system $\cF$. {\em Orthogonal
  sets} of $\cF$ are sets that do not overlap with any set in $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system
$\cF$. \tnote{define substitutio decomp.} This is later used to define
a certificate if $\cF$ has no \COP. \PQRtree is a data structure
defined in \cite{mm96} as a data structure to represent any set system
$\cF$ with additional information in their $R$-nodes if $\cF$ has no
\COP.  All three of these data structures are proposed as
generalizations of \cite{bl76}'s \PQtree and hence produce the
\PQtree if $\cF$ has \COP. As a whole, all these three data
strutures are near-identical with their differences being notional.
In this section, we will discuss the basic theory that they all hold
and stick to the name of the data structure as \PQRtree.

@@

An important observation made by \cite{mm96} is that if $\cF$ is a set
system with \COP then, aside from every set in $\cF$ being consecutive
(after applying the \COP order) the following sets must also be
consecutive for any $A, B \in \cF$.
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and $A \nsubseteq B$ respectively.
\end{enumerate}
Also note that trivially, sets in $\cT(U)$ are consecutive in any
permutation of $U$\footnote{$\emptyset$ is considered consecutive by
  convention.}.

@@

% \tnote[FYI]{ The results in cite~{bl76} on COT are based on the
%    result that interval graphs are AT-free chordal graphs cite~{lb62}?
%  No they don't seem related. this is the stupid bl lb confusion.}%
